Individual Partisan Harm and Utility in Redistricting

### 2021-10-14 paper/paper.Rmd
intro

Districting is a complicated process.
Residents of a state or locality must be partitioned into districts based on the geographies that they live in.
These districts must adhere to many constraints, often requiring components such as contiguity and equal population.
In drawing the districts, it's easy to lose sight of the individuals impacted by district drawing.
Often, we consider how parties succeed or fail under the electoral environments generated by enacted plans.
Yet this misses that parties are formed by voters who ally with parties.
Rather than considering parties, we ask how individuals are harmed or benefited by the districting process?

Decades of literature on redistricting have focused on the fairness to parties.
This literature defers to symmetric outcomes for parties, commonly referred to as partisan symmetry [@katz2020].
Partisan symmetry offers the idea that the number of seats awarded to a party by a set of votes would be the same if the labels on the parties were swapped.
Partisan symmetry is a useful quantity of interest which has been widely used to improve academic and legal understanding of the redistricting process.
We do not challenge the value of partisan symmetry, but instead suggest that a different question should be asked to describe the fairness of district based democracies.

We argue that understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
Our metrics offer a framework for understanding approximately individual harms, benefits, and net fairness.
Using counterfactual plans to understand this space allows us to properly calibrate understandings of the impact of redistricting plans on individual understanding.
Thus, we create measures which capture the possibility that no plans exist which favor one community [@duch:18].
Using this class of metrics which describe the individual affect via representation consequently allow us to better align legalistic interpretations of harm for standing with political science understandings.

To describe a reasonable counterfactual to measure our quantity of interest, we rely on advances in redistricting simulation methodologies.
Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly [@ciri:darl:orou:00; @chen:rodd:13; @deford2019]. These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations. Yet, without a known target distribution, these heuristic approaches to simulation fall do not offer necessary guarantees for confidence in the simulated plans to represent the total space of possible plans [@fifield2020essential]. Recent advances in Monte Carlo methods allow us move beyond simulation towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020].

Using these sampling methods, we can then develop a reliable counterfactual that can be tuned to consider the sampling constraints imposed by the laws and practices of a state.
This counterfactual allows us to characterize the likely plans from a state.
Often, we can would then call this an ensemble analysis, which is a flexible approach to sampling redistricting plans by simulation, relying on large numbers of plans for stability [@matt:hers:21].
using this counterfactual, we thus create measures of harm, utility, and fairness which are based on a single plan compared to a set of other reasonable and feasible plans which incorporate the necessary legal constraints to make a plan viable.

We offer a reminder that geographic context is key to understanding common problems in political science.
Elections themselves are geographically determined.
Recent research offers important resolution of the degree of partisan sorting in the US [@enos:brown:21].
By combining those understandings with the necessities of redistricting, we are able to expand the localized aspect of harm.
Further, we suggest that while the efficiency gap may not incorporate geographic information, it often helps capture individual harm better than other common measures.
This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.
We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.

TODO: Add findings!

```{=html}
<!--
Understanding fairness in electoral systems has long suffered from a focus on parties over the individuals who vote for them.
There is no doubt that parties are interesting units in electoral discussions.
They offer stability over candidates, information for the under-informed, and shortcuts for understanding preferred vote choices.
Yet, court histories and academic traditions value the harm of the individual over the harm to a transient political body, such as the party.

@@

Section \ref{sec:voter-util} provides the necessary notation and introduces our measures of voter utility.
Section \ref{sec:indiv-harm} builds a measure of individual harm and a derived measure of partisan fairness from this notion of utility.
Section \ref{sec:estimating} discusses estimating these measures from observed election results.
Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
Section \ref{sec:empirical} applies the proposed measures to a real-world example to illustrate their value and these comparsions.
Section \ref{sec:discussion} concludes.

@@

Partisan bias is probably the most widely-used measure of partisan symmetry.
The bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.

@@

Our measures greatly differ from the Mean-Median Difference.
With the focus on individuals, the considerations of the mean district are only related to individuals in-so-far as they describe the general propensity to be harmed or benefited by a particular map.
The consideration of a median district is similarly distanced.
As with partisan bias, as our measures do not attempt to measure partisan symmetry, our measures can be applied to congressional delegations.

>>>>

How do you measure a partisan gerrymander?
Researchers, courts, and the public may agree that gerrymandering involves the manipulation of district boundaries for political gain, but beyond this vague definition are liable to disagree on a particular operationalization.

Decades of academic literature on redistricting have focused on the fairness to parties.
The normative standard for redistricting in this line of research is for electoral outcomes to be symmetric across political parties---what is commonly referred to as *partisan symmetry* [@katz2020].
A variety of theoretical measures have been developed to capture this notion and to measure quantitatively any deviations from it.
These measures may generally (and, as some have argued, necessarily) be derived from the *seats-votes curve*, which relates the expected number of seats a party wins to its share of votes received.

The courts, meanwhile, have focused more on the impact an alleged gerrymander has on individual voters.
To proceed with a legal challenge to a gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the gerrymander [@1964reynolds; @1962baker].

Crucially, for a voter to be harmed by a districting plan, there must exist another districting plan which remedies the harm.
An example we will return to frequently is that of a Massachusetts Republican: based on residential patterns of party voters, it is not possible to draw a Republican congressional district in Massachusetts, and so a Republican voter cannot be harmed by a districting plan that yields a fully Democratic delegation [@duch:18].

We argue that understanding the redistricting process requires a feasible and reliable understanding of a counterfactual set of redistricting plans.
To that end, we define two individual utility measures for districting plans: a *local utility* capturing the value to a voter of having a representative of the same party, and a *global utility* reflecting the value in maximizing the number of seats awarded to a voter's chosen party.
From these two utility measures, we derive a measure of harm, based on counterfactual plans, and a notion of individual partisan fairness, which occurs when harms are evenly distributed across voters of each party.
Deviations from fairness are measured by *differential harm*, which, as we show, is intimately connected to the number of seats won by each party.

Taken together, our measures offer a framework for understanding the individual impacts of a districting plan, and how these impacts are borne by members of each party.
The use of counterfactual plans allows us to better relate legalistic interpretations of harm to those of political science.

To estimate these measures requires generating counterfactual districting plans, for which we rely on advances in redistricting simulation methodology.
Early methodologies for simulating plans relied on randomness to generate many different districting plans [@ciri:darl:orou:00; @chen:rodd:13; @deford2019]. These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations. Yet, lacking a specific target probability distribution, these heuristic approaches to simulation do not allow for probabilistic interpretations, and risk misrepresenting the total space of possible districting plans [@fifield2020essential]. Recent advances in Monte Carlo methods allow us move beyond random simulation and towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020]. We can thus develop a reliable set of counterfactual plans that can be tuned to reflect the constraints imposed by the laws and practices of a particular state [@matt:hers:21].

We offer a reminder that geographic context is key to understanding common problems in political science; elections themselves are geographically determined.
Recent research offers important resolution of the degree of partisan sorting in the US [@enos:brown:21].
By combining those understandings with the necessities of redistricting, we are able to expand the localized aspect of harm.
Further, we suggest that while the efficiency gap may not incorporate geographic information, it often helps capture individual harm better than other common measures.
This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.
We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.

```{=html}
<!--
Districting is a complicated process. Residents of a state or locality must be partitioned into districts based on the geographies that they live in. These districts must adhere to many constraints, often requiring components such as contiguity and equal population. In drawing the districts, it's easy to lose sight of the individuals impacted by district drawing. Often, we consider how parties succeed or fail under the electoral environments generated by enacted plans. Yet this misses that parties are formed by voters who ally with parties. Rather than considering parties, we ask how individuals are harmed or benefited by the districting process?

Partisan symmetry is a useful quantity of interest, which has been widely used to improve academic and legal understanding of the redistricting process. We do not challenge the value of partisan symmetry, but instead suggest that a different question should be asked to describe the fairness of district based democracies.

Understanding fairness in electoral systems has long suffered from a focus on parties over the individuals who vote for them.
There is no doubt that parties are interesting units in electoral discussions.
They offer stability over candidates, information for the under-informed, and shortcuts for understanding preferred vote choices.
Yet, court histories and academic traditions value the harm of the individual over the harm to a transient political body, such as the party.

@@

Section \ref{sec:voter-util} provides the necessary notation and introduces our measures of voter utility.
Section \ref{sec:indiv-harm} builds a measure of individual harm and a derived measure of partisan fairness from this notion of utility.
Section \ref{sec:estimating} discusses estimating these measures from observed election results.
Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
Section \ref{sec:empirical} applies the proposed measures to congressional redistricting in the state of New Jersey to illustrate their value and these comparisons.
Section \ref{sec:discussion} concludes.

@@

Partisan bias is probably the most widely-used measure of partisan symmetry.
The bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.
Each value of $s$ yields information about a different point on the seats-votes curve.

@@

The mean-median difference straightforwardly measures the difference between the mean and median district vote shares.
In practice, the mean-median differences is moderately correlated with other partisan redistricting measures.
However, considerations of the mean and median district are only related to individuals insofar as they describe the general propensity to be harmed or benefited by a particular map.
As [@katz2020] discuss, the mean-median difference can be interpreted as an estimator of *votes-denominated bias*, another value derived from the seats-votes curve.
